Accessible, realistic genome simulation with selection using \stdpopsim

### 2025-03-19 analysis2_manuscript.tex
one more tweak
Update analysis2_manuscript.tex
Co-authored-by: Andrew Kern <adkern@uoregon.edu>

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community 
    \citep[e.g.,][]{speidel2021inferring, wang2021automatic, yang2022trustgwas, degiorgio2022spatially, browning2023fast, schweiger2023ultrafast, temple2024modeling, haag2025pandora}.
    Yet, previous releases of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation \citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps 
    \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, 
    and the impact of background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a simple, unified framework for simulating both neutral and
    non-neutral evolutionary processes and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore, the framework should include the complex realities of 
    genomes, including heterogeneous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.
    %
    \stdpopsim{} provides a framework for such models,
    thus lowering the barrier to production of complex, multifaceted simulations,
    and allowing a wider range of researchers to make use of high-quality simulations
    in developing or testing their work.
    The project also promotes open, collaborative, and inclusive work,
    viewing capacity-building across the community as equal in importance to software development.

@@

    To test methods for detecting selective sweeps, we considered 100 evenly distributed 5cM
    segments of human chromosome 1.
    As recombination rate varies across the genome, the size of the region in base pairs simulated varied, however the recombination distance,
    which is critical for the linked effects of a sweep, was held constant.
    %IG: moved this up here since it makes mroe sense to describe it as if these segments were never simulated
    % also converted to rec rate from cM. Please check!
    Seven segments in which the central 1Mb had an average recombination rate below $10^{-11}$ were excluded from the analysis.
    Each segment was simulated in each of the following three settings:
    (1) neutral, (2) background selection (BGS), and (3) hard sweep with BGS.
    In all three settings, we applied a genetic map from the HapMap Project
    \citep[\stdpopsim label \texttt{HapMapII\_GRCh38}]{international2007second} and
    the three-population out of Africa (OOA) demographic model from
    \citet[\stdpopsim label \texttt{OutOfAfrica\_3G09}]{gutenkunst2009inferring}.
    In the two settings with BGS, we applied the DFE model with gamma-distributed selection coefficients inferred by
    \citet[\stdpopsim label \texttt{Gamma\_K17}]{kim2017inference} to exons annotated by the HAVANA group release 104
    \citep[\stdpopsim label \texttt{ensembl\_havana\_104\_exons}]{ensembl2018}.
    Hard sweeps were modeled by generating a single beneficial mutation 
    with selection coefficient of $s = 0.03$ in the center of the 5cM genomic segment
    and discarding simulations in which the beneficial mutation did not reach (or exceed) a terminal frequency of $95\%$.
    For each of the 100 genomic segments of chromosome 1, we generated 1,000 replicate datasets under each of the three settings.
    We used the \slim engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2.
    Each simulation generated ten diploid genomic segments for each of the two sampled populations: YRI and CEU.
    The sequences simulated with a hard sweep and BGS were used for sweep detection,
    and sequences simulated under the other two settings were used to calibrate the thresholds of the test statistics
    used by the three sweep-detection methods (see below).

>>>>

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community 
    \citep[e.g.,][]{speidel2021inferring, wang2021automatic, yang2022trustgwas, degiorgio2022spatially, browning2023fast, schweiger2023ultrafast, temple2024modeling, haag2025pandora}.
    Yet, previous releases of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation \citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps 
    \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, 
    and the impact of background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a simple, unified framework for simulating both neutral and
    non-neutral evolutionary processes and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore, the framework should include the complex realities of 
    genomes, including heterogeneous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.
    %
    \stdpopsim{} provides a framework for such models,
    thus lowering the barrier to production of complex, multifaceted simulations,
    and allowing a wider range of researchers to make use of high-quality simulations
    in developing or testing their work.
    The project also holds as a central goal open, collaborative, and inclusive work,
    viewing capacity-building across the community as equal in importance to software development.

@@

    To test methods for detecting selective sweeps, we considered 100 evenly distributed 5cM
    segments of human chromosome 1.
    As recombination rate varies across the genome, the size of the region in base pairs simulated varied, however the recombination distance,
    which is critical for the linked effects of a sweep, was held constant.
    %IG: moved this up here since it makes mroe sense to describe it as if these segments were never simulated
    % also converted to rec rate from cM. Please check!
    Seven segments in which the central 1Mb had an average recombination rate below $10^{-11}$ were excluded from the analysis.
    Each segment was simulated in each of the following three settings:
    (1) neutral, (2) background selection (BGS), and (3) hard sweep with BGS.
    In all three settings, we applied a genetic map from the HapMap Project
    \citep[\stdpopsim label \texttt{HapMapII\_GRCh38}]{international2007second} and
    the three-population out of Africa (OOA) demographic model from
    \citet[\stdpopsim label \texttt{OutOfAfrica\_3G09}]{gutenkunst2009inferring}.
    In the two settings with BGS, we applied the DFE model with gamma-distributed selection coefficients inferred by
    \citet[\stdpopsim label \texttt{Gamma\_K17}]{kim2017inference} to exons annotated by the HAVANA group release 104
    \citep[\stdpopsim label \texttt{ensembl\_havana\_104\_exons}]{ensembl2018}.
    Hard sweeps were modeled by generating a single beneficial mutation 
    with selection coefficient of $s = 0.03$ in the center of the 5cM genomic segment
    and discarding simulations in which the beneficial mutation did not reach (or exceed) a terminal frequency of $95\%$.
    For each of the 93 genomic segments of chromosome 1, we generated 1,000 replicate datasets under each of the three settings.
    We used the \slim engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2.
    Each simulation generated ten diploid genomic segments for each of the two sampled populations: YRI and CEU.
    The sequences simulated with a hard sweep and BGS were used for sweep detection,
    and sequences simulated under the other two settings were used to calibrate the thresholds of the test statistics
    used by the three sweep-detection methods (see below).
